---
title: "MicroEcoTools"
author: "Soheil A. Neshat"
date: "2025-03-10"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{MicroEcoTools Vignette}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options: 
  markdown: 
    wrap: 72
---

```{r setup, include=FALSE}
# Set global options for code chunks
knitr::opts_chunk$set(echo = TRUE, warning = TRUE, message = TRUE)
# Load the MicroEcoTools package
#devtools::load_all()
```

## Introduction

MicroEcoTools is an R package designed to provide theoretical tools for
microbial community analysis. The package includes functions for:

Comparing traits or taxa across experimental groups using Welch-ANOVA.
Performing pairwise comparisons and assigning CSR (Competitor–Stress
tolerant–Ruderal) categories. Simulating community data to assess the
probability of CSR assignments. Calculating diversity metrics such as
Hill diversity. In this vignette, we demonstrate how to prepare your
data (including converting SummarizedExperiment or phyloseq objects) and
run the core analyses provided by the package.

This package can be used on experimental community data to investigate the 
relative contribution of deterministic and stochastic assembly mechanisms through
a taxa based null model analysis. In addition, Grime's CSR categories can be 
assigned to taxa and traits in communities under different level of disturbance.

In this tutorial, we a use a dataset obtained through a perturbation experiment:

## Installation

This chunk installs the MicroEcoTools package.

```{r installation}
if (!requireNamespace("devtools", quietly = TRUE))
  install.packages("devtools")
devtools::install_github("Soheil-A-Neshat/MicroEcoTools")
library(MicroEcoTools)
```

## Data prepration

MicroEcoTools functions require input data in a data frame format. The
expected format is:

Column 1: Experimental group (e.g., Exp.Grp) Column 2: Replicate (e.g.,
Replicate) Remaining columns: Count or abundance data for taxa or traits

If your data are in a SummarizedExperiment or phyloseq object, the
functions automatically check the input and call the helper function
?convert_to_microecotools_df to convert it
into the required format.

## Example dataset

For demonstration, we have included three datasets from the following references:
1. Santillan, Ezequiel, et al. "Trait‐based life‐history strategies explain succession scenario for complex bacterial communities under varying disturbance." Environmental microbiology 21.10 (2019): 3751-3764.
2. Santillan, Ezequiel, et al. "Frequency of disturbance alters diversity, function, and underlying assembly mechanisms of complex bacterial communities." npj Biofilms and Microbiomes 5.1 (2019): 8.

Explanation for the study, data, etc ....

```{r CSR IP2G data, echo=TRUE}
dim(CSR_IP2G_data)
head(CSR_IP2G_data)
```

```{r CSR IP2G data, echo=TRUE}
dim(CSR_TAXA_data)
head(CSR_TAXA_data)
```

```{r Null model analysis data, echo=TRUE}
dim(NMA_data)
head(NMA_data)
```



##CSR Assignment CSR_assign function analyses the dataset and assigns
competitive, stress-tolerant and ruderal categories to traits/taxa which
are abundant in undisturbed, highly disturbed and intermediately
disturbed communities. The output for this function is a list containing
comparison tables and plots.

```{r CSR assignment, echo=FALSE}
# Perform CSR category assignment
csr_result <- CSR_assign(dAtA = CSR_IP2G_data, var.name = "Trait", p_adj = "BH",
                         v.equal = FALSE, p.value.cutoff = 0.05, Parallel = TRUE, Verbose = FALSE)
summary(csr_result)
```

```{r CSR plotting, echo=FALSE}
# Plot the CSR plot:
# Traits sorted based on abundance 
CSR_plot(CSR_results = csr_result, plot_type = "abundance")
# Traits sorted based on p-value
CSR_plot(CSR_results = csr_result, plot_type = "p_value")
```

```{r CSR assignment summary, echo=FALSE}
# Summary of the CSR assignment can be seen using:
summary(csr_result)
```

##CSR Assignment with simulation CSR_simulation function generates
simulated communities with similar multinomial distribution to the
observed communities and perfroms CSR category assignment. The
difference between this function and CSR_assign is that this function
calculates a probability value which shows accuracy of the category
assignment.

```{r CSR assignment with simulation, echo=FALSE}
# Run CSR_Simulation on the dataset.
simulation_result <- CSR_Simulation(DaTa = CSR_IP2G_data, NSim = 2, p_adj = "BH",
                                    var.name = "Trait", v.equal = FALSE, p.value.cutoff = 0.05,
                                    Parallel = TRUE, NuLl.test = FALSE, Keep_data = FALSE)
summary(simulation_result)
```

##Welch-ANOVA Analysis The Welch_ANOVA function compares traits or taxa
across experimental groups using Welch's ANOVA. It automatically
switches to an equal variance test if needed.

```{r Welch-Anova, echo=FALSE}
# Run Welch_ANOVA on the example dataset.
result_welch <- Welch_ANOVA(dAtA = CSR_IP2G_data, var.name = "Trait", p_adj = "BH", Parallel = FALSE)
# Print the resulting comparison table
summary(result_welch)
```

##Pairwise Welch-Anova This function performs a pairwise comparison for
taxa/traits between experimental groups.

```{r Pairwise Welch-Anova, echo=FALSE}
# Generate a pairwise comparison table using pairwise_welch.
pairwise_result <- pairwise_welch(dAtA = CSR_IP2G_data, var.name = "Trait", p_adj = "BH",
                                  v.equal = FALSE, p.value.cutoff = 0.05, Parallel = TRUE)
summary(pairwise_result)
```

##Null model analysis We can assess the relative contribution of
stochastic and deterministic assembly mechanisms in a community using
the NMA function. Running the following code chunk does this analysis.
This function takes in the observed communities, pool all the taxa and
generate random communities from the pool. Then, it compares the
observed communities and randomly generated communities to calculate an
effect size. The higher the effect size the more deterministic the
community assembly.

```{r Null model analysis, echo=FALSE}
nma_results <- NMA(DaTa = NMA_data, NSim = 10, InDeX = "invsimpson")
summary(nma_results)
#you can look at the standard effect size plot generated using richness as the measure of diversity using
NMA_plot(nma_results, plot_type = "SES_species_number")
#To see the plot generated for standard effect size using the specified diversity index use
NMA_plot(nma_results, plot_type = "SES")
```
